Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center

### 2024-02-28 paper/markdown/review_response.Rmd

\newpage

# Editor

Dear Editor,

Thank you very much for the positive and very thoughtful review of our manuscript. We appreciate a lot the reviewers’ comments and suggestions to improve our manuscript. 
We incorporated them in the revised text.
Please find below the point-to-point response to the reviewers’ issues. 
Since the reviewers requested more information and literature evidence, we also deleted few less important references from the initial submission to keep the bibliography within reasonable limits. 
We also introduced three major changes to the manuscript text and data independently of the reviewers' requests:

1. We re-classified accident sport types according to the official scheme of the Austrian Board of Mountain Safety [@OsterreichischesKuratoriumfurAlpineSicherheit2024]. This allowed us to compare the profile of accident sports in our cohort with the national registry data. Please note, that the re-classification of the sport types requires us to re-train machine learning models of the mental health cluster assignment. For this reason, minor differences in performance of the models were observed as compared with the initial submission (e.g. in __Figure 7__). Still, they did not affected the general conclusions.

2. We identified clinically relevant symptoms of anxiety and depression with the uniform $\geq$ 10 point cutoff of, respectively, the GAD-7 and PHQ-9 tools in accordance with the work by Humer et al. This change allowed us for a direct comparison of frequency of clinically relevant anxious and depressive symptoms in our manuscript and the estimates for the general Austrian population at the time of our study survey [@Humer2022].

3. We provided additional numeric measures of quality of the mental clusters, misclassification rate [@Rousseeuw1987] and neighborhood preservation [@Venna2001], which were used by us to argument for the excellent reproducibility of the mental clusters (__Supplementary Figure S8__).

Additionally, some minor copyediting was done. 
Changes in the main manuscript text are highlighted in red. 
We hope that the manuscript is now suitable for publication in _European Archives of Psychiatry and Clinical Neuroscience_.

@@

## Point 2

__Abstract__: The abstract is well structured. 
Probably it might be better not to make the first sentence about a topic which the study does not cover (positive effects of mountain sports).

## Response 2

We agree that the first sentence of __Abstract__ does not specifically relate to the subject of our research, which concerns primarily negative psycho-pathological burden of mountain sport accidents.
We changed this sentence of __Abstract__ attempting to accommodate the reviewer's concern but did not completely remove it because we feel that especially “mountain sports” are generally believed to bolster mental health. 
Additionally, individuals performing mountain sports are often portrayed as resilient personalities protected from mental disorders, which we can in part support with our data (__Supplementary Figures S4__ and __S5__). 
In our clinical practice, these beliefs are a common source of stigma in members of the mountaineering community. 
So I hope this makes it more clear why we did not remove the first sentence completely and hopefully the reviewer can understand our intentions.

@@

## Point 3

__Introduction__: The introduction is well written. 
At the end I think it would help to make the hypothesis more clear. 
I know that with SVM methods a hypothesis free approach is common, however the authors intend to (1) categorise the sample (2) find predictors.

## Response 3

Thank you for this suggestion. 
In the last paragraph of __Introduction__, we stressed the exploratory character of the study and the general strategy, i.e. classification of participants by their mental health measures followed by a search for predictors of mental health classes/clusters.
We also re-wrote __Methods/Analysis endpoints accordingly__.

@@

We absolutely agree with your comment that the low response rate is a source of bias. 
Unfortunately this is a challenge of many studies tackling with mental disorders such as PTSD, anxiety and depression.

To assess the potential effects of the low response rate on the cohort composition we included an extensive comparison of the analysis cohort with invited individuals and participants excluded due to incomplete psychometric data in the initial submission. 
In the revised manuscript we investigated the potential selection bias in more detail, by comparing sociodemographic features of the cohort with the general Austrian population and by comparing characteristic of the mountain sport accident in the study cohort with a nation-wide collective of Austrian mountain accident victims.

As presented in __Supplementary Table S6__ and discussed in __Results/Characteristic of the study cohort__, the study sample was characterized by significantly higher frequencies of males, persons with tertiary education grade and actively employed as compared with the general Austrian population. 
In turn, rates of smokers and individuals suffering from chronic physical or mental disorders was significantly lower. 
Effect sizes of those differences were, however, very small (Cramer's V 0.00078 to 0.0025). 
Results of the comparison of the accident characteristic are presented in __Supplementary Figure 1__ and described in __Results/Characteristic of the study cohort__. 
They indicate significantly higher frequency of winter sport accidents in the study cohorts than in the Austrian data set. 
Yet the effect size of the difference was again small (Cramer’s V ≤ 0.14). 
Taken together, these analyses suggest that there is indeed a weak but significant selection bias towards men, well educated, professionally active and healthy individuals, and victims of winter sport accidents. 
We point this out in the limitation section of __Discussion__ of the revised manuscript.

To address the possible under- or overestimation (see: __Point 10__, __Reviewer 2__) of mental disorder symptoms in the study cohort, we compared frequency of traumatic events, manifest PTSD, symptoms of anxiety and depression in the study cohort with large survey studies [@Hauffa2011; @Darves-Bornoz2008; @Kilpatrick2013; @Koenen2017; @Kessler2017; @Humer2022], as well as an Austrian 2019 general health microcensus study [@StatistikAustria2019a]. 
Frequency of traumatic events and manifest PTSD in our sample was lower or comparable with the estimated prevalence in the general German population reported by Hauffa and colleagues [@Hauffa2011] (__Supplementary Figure S4A__). 
Interestingly, frequency of manifest PTSD and frequency of the PCL-5 domain B, C, D, and E PTSD symptoms did not differ significantly from figures published by Mikutta et al. for Swiss mountain rescuers [@Mikutta2022] (__Supplementary Figure S4B__). 
Furthermore, frequency of clinically relevant anxious (`r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'anxiety')`) and depressive symptoms (`r lit_plots$depr_anx_stats %>% get_est_ci(source == 'cohort', disorder == 'depression')`) assessed, respectively by the GAD-7 and PHQ-9 instruments, was significantly lower in the study cohort as compared with the 2022 estimates for the general Austrian population reported by Humer et al. (anxiety: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'anxiety')`, depression: `r lit_plots$depr_anx_stats %>% get_est_ci(source == 'humer_2022', disorder == 'depression')`, __Supplementary Figure S5A__) [@Humer2022]. 
These results let us reason that, despite the mountain accident, the overall mental health status of the study cohort was comparable to or better than the general population of German-speaking countries and we elaborate on this in __Results/Characteristic of the study cohort__ and __Discussion__.

@@

Throughout the discussion: Please find a clear terminology of the PTSD: sometimes you talk about manifest PTSD sometimes about PTSD symptoms. 
As you mostly do refer to symptoms of PTSD I’d stick with that.

## Response 7

Thank you for raising this important issue. 
We agree that the distinction between symptoms of PTSD and manifest PTSD was not completely coherent in the initial text. 
We define symptoms of PTSD as positivity for at least one domain of the PCL-5 tool. 
Manifest PTSD is considered for participants screened positive for all four domains of the PCL-5 instrument [@Bovin2016]. 
This distinction is described now in __Methods__ and __Supplementary Methods__.

Of note, while single PTSD symptoms affected nearly one of five study participants, manifest PTSD was discerned in four individuals only. 
In the revised text, we defined this distinction in __Methods__ and __Supplementary Methods__ and use consistently the terms 'manifest PTSD' and 'symptoms of PTSD' throughout the manuscript.

@@

## Point 10

I would almost certainly expect there to be a strong self-selection bias in the obtained responses to your survey (i.e., those still dealing with the consequences of their accident), likely overestimating the prevalence of mental health-related impairment in these individuals. 
In addition, the supplementary table S4 shows that respondents had more severe injuries requiring hospitalization and surgery compared to non-respondents, also reflecting self-selection of particularly affected individuals among those who experienced a mountaineering accident. Does that change the implication(s) of your results?

## Response 10

We absolutely agree that there is a selection bias in this study. 
This was already pointed out in the limitations section of __Discussion__ of the initial submission. 
As described in more detail in __Response 4__ we took now additional measures to better characterize the potential selection bias. 
From comparisons of our sample with the general Austrian population (__Supplementary Table S4__) and the collective of mountain accident victims in Austria in 2023 (__Supplementary Figure S1__), we infer that the study cohort consisted of significantly more males, highly educated and healthy individuals, and victims of winter sport accidents that expected for the general population. 
Effect size of those differences was weak (Cramer's V effect size statistic $\leq$ 0.3)
We elaborate on those findings in __Results/Characteristic of the study cohort__ and summarize these findings on the selection bias in the limitation part of __Discussion__. 
Of note, the selection bias may explain low frequency of traumatic events prior to the mountain accident and rare cases of manifest PTSD in our cohort, as e.g. men and tertiary school graduates are less likely to be exposed to traumatic events and develop PTSD in the lifetime [@Darves-Bornoz2008] (__Supplementary Figure S1__). 
We describe this possibility in __Discussion__ of the revised manuscript.

Next, we addressed the issue of under- or over-reporting of mental health symptoms in the study sample, which may be attributed to the selection bias of the sample. 
To this end, we compared frequency of traumatic events, manifest PTSD, anxious and depressive symptoms in the study cohort with estimates for the general population (__Supplementary Figure S4__). 
As outlined in __Response 4__ in more detail, frequency of traumatic events and manifest PTSD was at the lower tail of the reported prevalence and comparable with figures for the German population published by Hauffa and colleagues [@Hauffa2011]. 
Anxious symptoms measured with the GAD-7 instrument were less common in the study cohort than in the Austrian population during the survey [@Humer2022]. 
In turn, depending on the tool and time point of survey, overall depressive symptoms and symptoms of moderate-to-severe depression were lower or comparable with the general Austrian population (__Supplementary Figure S5__). 
We discuss in __Results/Characteristic of the study cohort__ and in __Discussion__, that the overall frequency of mental disorder symptoms in our cohort is, in spite of the accident, comparable with or event lower than in the general population. 
It is hence possible that bias of the sample towards protective characteristic (education, male sex, economic status) compensates the negative effects of the accident and injury on mental health. 
Additionally, in light of the comparison of the mental disorder symptoms in the cohort and estimates for the general population, over-reporting of mental disorder symptoms in the study cohort is unlikely.

@@

rev_figures$time$plot

```

__Review Figure R\@ref(fig:fig-time). Effects of ward admission-survey time and of accident year on the major psychometric readouts and the mental cluster assignment.__

_(A) Differences in the major scores of mental health and quality of life as a function of the ward admission - survey time (PTSD: PCL-5 score, sum of scores for the B, C, D, and E domains; post-traumatic growth: PTGI, sum of all dimensions; anxiety: GAD-7, depression: PHQ-9; resilient coping: RS13; quality of life: EUROHIS-QOL 8, arithmetic mean of the domain scores). Statistical significance was determined by Kruskal-Wallis test with $\eta^2$ effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Score values are presented in violin plots. Single observations are visualized as points. Red diamonds with whiskers represent medians with interquartile ranges. Effect sizes and p values are displayed in the plot captions. Numbers of complete observations are displayed in the X axes._

@@

## Point 13

According to your statement on page 13 (lines 4-9), it seems that your sample was especially resilient compared to the general population. 
Wouldn't this perhaps imply a protective factor of (mountain) sports overall, despite accident risk?

## Response 13

Absolutely! 
We make this more clear now in __Discussion__ section to emphasize this point. 
To support this notion with hard figures, we compared average scores with samples from the general German [@Leppert2008] and Tyrolean/Austrian population [@Chernova2021], and with resilience scores published by Mikutta and colleagues for a cohort of Swiss mountain rescue service workers [@Mikutta2022]. 
Supporting the common belief of positive effects of mountain activity on mental strength, resilience scoring of our cohort (despite the accident) and of the Swiss mountain rescue sample (despite high prevalence of traumatic events) was substantially higher than in the general population (__Supplementary Figure S5__). 
This may at least in part explain low frequency of manifest PTSD and PTSD symptoms in those two collectives (__Supplementary Figure S4B__). 
We elaborate on this new data in more detail in __Results/Characteristic of the study cohort__ and in __Discussion__.

@@

## Point 15

What new insights does this study offer? 
Pre-existing mental disorders and persistent somatic symptoms are well-known risk factors for the development of psychiatric disorders. 
Indeed, a reliable model identifying at-risk individuals would have been interesting, but this study could not provide one.

## Response 15

Alone in Austria with 9 million inhabitants, there are over 900000 members of diverse alpine clubs [@VAVOE2024], which amounts to an enormous community of mountain sport enthusiasts, professional and free-time athletes. 
The dark side of popularity of mountain sports are accidents whose number exceeded `r table(lit_kurasi$data$sport_detail$subset)['Austria'] %>% signif(2)` in 2023 in the nation-wide scale, according to the official counts by the mountain safety board [@OsterreichischesKuratoriumfurAlpineSicherheit2024]. 
By comparison, this number corresponds approximately to the cumulative yearly incidence of prostate and breast cancer in Austria [@StatistikAustria2023a]. 
Despite these huge figures and the burden of mountain sport accidents on health system and economy of the alpine region, only a handful of papers has been published, which tackle single, isolated aspects of mental health of mountain sportspersons and victims of mountain sport accidents [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004]. 
Our study provides an integrated and 'holistic' characteristic of mental health after a mountain sport accident, by assessment of multiple dimensions of mental readouts with semi-supervised clustering and machine learning. 
This is the prime point of novelty of our work.  
Despite the positive effects of mountain sports on physical and mental health and the common perception of sportspersons as mentally strong individuals, our data suggest that a large subset of mountain accident victims, the PTS cluster, is prone to long-term mental health deterioration. 
This is another novel finding of out work. 
As pointed out by the reviewer, there are many reports on markers and diagnostic tools of post-traumatic stress [@Sandweiss2011; @Shalev2019; @Schultebraucks2020; @Ziobrowski2021]. 
Yet, we were not able to establish a reliable set of predictors of mental health disorders after the mountain sport accident. 
This negative finding, which has not been reported so far, has profound clinical implications and underscores the need of widely accessible psychological support in management of this specific accidents. 
We stress there three novel findings of our work in the first paragraph of __Discussion__ and the conclusion part of the revised manuscript.

@@

## Point 17

Why do you think did 20% of respondents not complete the survey? 
Why did they send back an incomplete battery?

## Response 17

This is a good point, thank you! 
As a part of exploratory analysis of the study data set, we routinely investigate missingness of variables. 
As presented in __Reviewer Figure R\@ref(fig:fig-miss-mental)__, the highest percentage of missing observations among the mental health assessment battery was observed for the resilience, depression, panic, quality of life, and anxiety variables. 
Those tools were missing in all or almost all of the `r table(miss_mental$analysis_tbl$status)['excluded']` respondents excluded from the analysis due to incompleteness of the psychometric data. 
In turn, for the PCL-5 instrument measuring PTSD symptoms, the missingness rates were the lowest (< 50% of excluded respondents). 
We can hence only speculate, that the excluded survey participants found questions of the questionnaire concerning their fears, concerns, life satisfaction, and mental health problems too private. 
This notion was also supported by results of the sensitivity analysis of the mental health clusters, where the majority of 'imputable' participants with missing psychometric data were assigned to the PTS cluster (__Supplementary Figure S11__).

@@

## Response 18

Thank you for raising this important point. 
Unfortunately, we can not answer it with published data. 
A query in the data base of mountain accidents in Austria (Austrian Board of Mountain Safety) did not help either, because the records are fully anonymized, and assignment of multiple accidents to the same victim is not possible.

Based on the comparison of the accident sports in the study cohort and the nation-wide collective of accident victims in 2023 (__Supplementary Figure S1__), we can conclude that the global difference in sport frequency distribution was significant but weak ($\chi^2$ test with Cramer's V effect size statistic: `r filter(lit_kurasi$test, variable == 'sport_detail')$plot_cap`). 
In particular, we could not observe higher percentages of sports generally perceived as risky such as ski touring, ice climbing, climbing, mountaineering or paragliding in the analyzed cohort as compared with the Austrian data set. 
This let us assume, that there are no large subsets of the cohort exposed to particularly high risk of injury.

Additionally, we did not ask about severity or time frame of mountain sport accidents in the past, so it is hard to judge if and how they affected mental health at the time point of the survey. 
As inferred from results of the sensitivity analysis presented in __Supplementary Figure S10__, prior mountain sport accidents had little impact on clustering of the study participants by their mental health features. 
They frequency also did not differ between the mental health clusters (__Supplementary Figure S12__). 
For these reasons it is unlikely that mountain accidents in the past had a substantial effect on the outcome of our study. 
We touch upon this point in __Results/Three clusters of mental health response after mountain sport accidents__ and __Results/Demographic, socioeconomic and clinical characteristic of the mental health clusters__.

### 2024-02-28 paper/markdown/manuscript.Rmd
Revised figures, tables, manuscript and supplement

__Background:__ The restorative effect of physical activity in natural alpine environments on mental and physical health is increasingly recognized. 
However, mountain sports are also linked to a risk of accidents. 
We aimed to characterize mental health in individuals following mountain sport accidents requiring professional medical management.

@@

Diagnostic symptoms of PTSD  include re-experiencing (criterion B), avoidance (criterion C), negative alterations in cognition and mood (criterion D) and hyper-arousal (criterion E) [@AmericanPsychiatricAssociation2013], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence. 
PTSD may co-occur with symptoms of depression and anxiety and substance abuse disorders [@Gagnon-Sanschagrin2022; @Brady2000]. 
Although most individuals have experienced at least one traumatic event during their lifetime [@Benjet2016], the prevalence of PTSD in the general European population has been estimated between 0.38 to 6.67% [@Burri2014]. 
Yet, up to 26% of acute physical trauma patients have been reported to develop PTSD [@Joseph2020].

So far it has proven difficult to reliably identify early predictors of PTSD development in accident victims. 
While some recent studies on specific populations have identified risk factors, such as pre-existing psychological distress [@Ziobrowski2021] or symptoms of mental disorders [@Papini2018], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020]. 
These results require validation and further generalization before they can be implemented into routine care [@Wafa2019; @Shalev2019].

Resilience, i.e. the ability to recover from adversities can protect an individual from developing PTSD and can help to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007; @Bensimon2012]. 
Generally individuals who perform mountain sports display high resilience values [@Ower2019]. 
Post-traumatic growth can occur in individuals struggling to overcome the consequences of a traumatic event and describes a psychological adjustment reaction with strengthened positive perceptions of self, others, one's life, and meaning of the event [@Tedeschi1996]. 
Although both resilience and post-traumatic growth are salutogenetic concepts their exact link is still under investigation [@Levine2009].

With the increasing popularity of mountain sports, it is essential to characterize the mental health consequences of accidents during mountain sports. 
Prevention of PTSD or other mental health problems following an accident is essential to preserve the positive mental health effects of physical activity in an mountain environment despite the accident risk. 
Available data on the consequences of accidents during mountain sports on mental health are scarce. 
Prior studies were performed on mountain guides, rescuers or avalanche accident survivors and mostly lack clinical data [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004]. 
We analyzed mental health readouts such as symptoms of PTSD, anxiety, depression, and somatization, resilience, sense of coherence, post-traumatic growth and quality of life in `r nrow(ptsd$dataset)` individuals treated at the Hospital of the Medical University of Innsbruck (Austria) after a mountain sport accident by means of semi-supervised clustering. 
Additionally, we searched for early sociodemographic and clinical predictors of mental health impairment following a mountain sport accident with a machine learning approach.

@@

Individuals treated following a mountain sport accident at the Department for Orthopedics and Traumatology at the University Clinic of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 were screened for participation. 
Individuals fulfilling the inclusion criteria: 
hospital admission at least 6 months prior to the start of the study, 
age $\geq$ 18 years, 
residence in a <span custom-style = 'revision'>German-speaking regions (Germany, Austria, Switzerland and South Tyrol/Alto Adige in Italy)</span>, and 
proficiency in German (n = `r nrow(ptsd$cleared)`) 
were invited to participate in the online study survey via conventional mail. 
Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey. 
Surveys of `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (overall response rate: `r signif(nrow(ptsd$dataset)/nrow(ptsd$cleared), 2) * 100`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__).

@@

The psychometric battery consisted of German versions of assessment tools for symptoms of 
PTSD (PCL-5: PTSD checklist for DSM-5) [@Bovin2016], 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ-9, <span custom-style = 'revision'>PHQ-2 and PHQ-8: 9-, 8- and 2-item patient health questionnaires</span> for depressive symptoms) [@Grafe2004], 
panic (PHQ-panic module) [@Grafe2004], 
somatization (PHQ-15: 15-item patient health questionnaire of common somatic symptoms as a substitute for somatization) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
and post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996]. 
<span custom-style = 'revision'>Items of the PTGI and PTSD instruments were provided with a captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your accident').</span>
The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).

Clinically significant symptoms of anxiety (GAD-7 $\geq$ 10), 
depression (PHQ-9 $\geq$ 11) [@Manea2012], 
somatization (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
resilience classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73) [@Leppert2008] 
were defined with published cutoffs.

@@

The primary analysis endpoint was assessment of mental health following a mountain sport accident by clustering using standardized psychometric tools for symptoms of PTSD, anxiety, depression, panic, somatization, resilience, sense of coherence, post-traumatic growth, and quality of life. 
The secondary analysis endpoints were 
(1) to compare demographic, socioeconomic and clinical characteristics between the different mental health clusters of mountain accident victims and 
(2) to identify patients at risk of mental disorders based on a set of explanatory variables available during acute medical treatment after mountain sport accidents.

@@

Surveys with complete psychometric data from `r nrow(ptsd$dataset)` individuals were analyzed (__Figure \@ref(fig:fig-consort)__). 
The median time between the trauma center admission after the accident and the survey completion was 
`r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and fewer surgery diagnoses (__Supplementary Table S4__). 
The survey participants excluded from the analysis due to incomplete psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__). 
Effect sizes of the differences between the included and excluded participants were weak (V $\leq$ 0.22 or r $\leq$ 0.15).

>>>>

__Background:__ <span custom-style = 'revision'>The restorative effect of physical activity in alpine environments on mental and physical health is recognized. However, risk of accidents and post-accident mental health problems is inherent to every sport, including mountain sport activity.</span>
We aimed to characterize mental health in individuals following mountain sport accidents requiring professional medical management.

@@

Diagnostic symptoms of PTSD  include 
re-experiencing (such as involuntary, intrusive memories, dreams or flashbacks of the traumatic event;, criterion B), 
avoidance (such as avoiding memories, thoughts or external reminders related to the traumatic event, criterion C), 
negative alterations in cognition and mood (criterion D) and
hyper-arousal (such as irritability, problems with sleep or concentration, criterion E) [@AmericanPsychiatricAssociation2013], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence. 
PTSD may co-occur with symptoms of depression and anxiety and substance <span custom-style = 'revision'>use</span> disorders [@Brady2000]. 
Although most individuals have experienced at least one traumatic event during their lifetime [@Benjet2016], the prevalence of PTSD in the general European population has been estimated between 0.38 to 6.67% [@Burri2014]. 
Yet, up to 26% of acute physical trauma patients have been reported to develop PTSD [@Joseph2020].

So far it has proven difficult to reliably identify early predictors of PTSD development in accident victims. 
While some recent studies on specific populations have identified risk factors, such as pre-existing psychological distress <span custom-style = 'revision'>or mental disorders</span> [@Ziobrowski2021], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020]. 
These results require validation and further generalization before they can be implemented into routine care [@Wafa2019; @Shalev2019]. 
<span custom-style = 'revision'>Overall, combination of multiple demographic, clinical and mental factors seem to yield more accurate PTSD predictions compared to single variables [@Papini2018].</span>

Resilience, i.e. the ability to recover from adversities can protect an individual from developing PTSD and can help to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007]. 
Generally individuals who perform mountain sports display high resilience values [@Ower2019]. 
Post-traumatic growth can occur in individuals struggling to overcome the consequences of a traumatic event and describes a psychological adjustment reaction with strengthened positive perceptions of self, others, one's life, and meaning of the event [@Tedeschi1996]. 
Although both resilience and post-traumatic growth are salutogenic concepts their exact link is still under investigation [@Levine2009].

With the increasing popularity of mountain sports, it is <span custom-style = 'revision'>key</span> to characterize the mental health consequences of accidents during mountain sports. 
Prevention of PTSD or other mental health problems following an accident is essential to preserve the positive mental health effects of physical activity in a mountain environment despite the <span custom-style = 'revision'>inherent</span> accident risk. 
Available data on the consequences of accidents during mountain sports on mental health are scarce. 
Prior studies were performed on mountain guides, rescuers or avalanche accident survivors, mostly lack clinical data<span custom-style = 'revision'>, and focus on single aspects of mental health such as trauma exposure or PTSD</span> [@Brugger2002; @Mikutta2022; @Traber2020; @Leonard2021; @Thordardottir2015; @Sommer2004]. 
<span custom-style = 'revision'>A characteristic of multiple facets of mental health following a mountain sport accident, including pathological and salutogenic features, is not clear at the moment.</span> 
<span custom-style = 'revision'>Additionally, risk factors of mental health problems specific for mountain sport accidents in a non-professional setting have not been reported so far.</span> 
<span custom-style = 'revision'>Identification of such risk-factors, and especially predictors available during acute medical management of the accident or a combination thereof, would greatly facilitate management of mountain sport accident by early, targeted psychological and psychiatric interventions.</span>

<span custom-style = 'revision'>Here, we aimed to identify long-term, persistent patterns of mental health in a sample of `r nrow(ptsd$dataset)` individuals treated at the hospital of the Medical University of Innsbruck (Austria) after a mountain sport accident. 
Such patterns were defined by symptoms of PTSD, anxiety, depression, and somatization, resilience, sense of coherence, post-traumatic growth and quality of life in a semi-supervised clustering analysis.</span>
<span custom-style = 'revision'>Subsequently, by statistical hypothesis testing and machine learning, we searched for early sociodemographic and clinical factors that characterize the newly defined mental health subsets of patients with a special focus on individuals prone to mental health impairment.</span>

@@

Individuals treated following a mountain sport accident at the Department for Orthopedics and Traumatology <span custom-style = 'revision'>of the hospital of the Medical University of Innsbruck</span> between 1^st^ January 2018 and 31^st^ December 2020 were screened for participation. 
Individuals fulfilling the inclusion criteria: 
hospital admission at least 6 months prior to the start of the study, 
age $\geq$ 18 years, 
residence in a <span custom-style = 'revision'>German-speaking regions (Germany, Austria, Switzerland and South Tyrol/Alto Adige in Italy)</span>, and 
proficiency in German (n = `r nrow(ptsd$cleared)`) 
were invited to participate in the online study survey via conventional mail. 
Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey. 
Surveys of `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (overall response rate: `r signif(nrow(ptsd$dataset)/nrow(ptsd$cleared), 2) * 100`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__).

@@

The psychometric battery consisted of German versions of assessment tools for symptoms of 
PTSD (PCL-5: PTSD checklist for DSM-5) [@Bovin2016], 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ-9, <span custom-style = 'revision'> and PHQ-8: 9- and 8-item patient health questionnaires</span> for depressive symptoms) [@Grafe2004; @Kroenke2002a], 
panic (PHQ-panic module) [@Grafe2004], 
somatization (PHQ-15: 15-item patient health questionnaire of common somatic symptoms as a substitute for somatization) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
and post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996]. 
<span custom-style = 'revision'>Items of the PTGI and PTSD instruments, and questions concerning flashbacks were provided with captions indicating that the questions refer to the mountain sport accident of interest ('The following questions refer to consequences of your mountain sport accident').</span>
The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).

Clinically significant symptoms of anxiety (GAD-7 $\geq$ 10), 
<span custom-style = 'revision'>depression (PHQ-9 $\geq$ 10) [@Humer2022],</span> 
somatization (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
resilience classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73) [@Leppert2008] 
were defined with published cutoffs.

@@

<span custom-style = 'revision'>The primary analysis endpoint was identification of subsets of patients differing in long-term symptoms of PTSD, anxiety, depression, panic, somatization, resilience, sense of coherence, post-traumatic growth, and quality of life following a mountain sport accident. This endpoint was pursued by semi-supervised clustering analysis.</span> 
<span custom-style = 'revision'>The secondary analysis endpoints were (1) comparison of demographic, socioeconomic and clinical characteristics of these newly defined subsets in a search for predictors of mental health disorders following a mountain sport accident, and (2) identification of</span> patients at risk of mental disorders based on a set of explanatory variables available during acute medical treatment after mountain sport accidents.
<span custom-style = 'revision'>The secondary analysis endpoints were addressed by statistical hypothesis testing and machine learning.</span>

@@

Surveys with complete psychometric data from `r nrow(ptsd$dataset)` individuals were analyzed (__Figure \@ref(fig:fig-consort)__). 
The median time between the trauma center admission after the accident and the survey completion was 
`r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents and lower rate of alpine ski/snowboard accidents (<span custom-style = 'revision'>significant, `r filter(excl$test$no_response, variable == 'sport_type')$plot_cap`</span>). 
<span custom-style = 'revision'>As compared with the analyzed participants participants, the non-responders were also characterized by significantly more frequent minor and less frequent severe injuries (`r filter(excl$test$no_response, variable == 'injury_sev_strata')$plot_cap`), overall lower injury grade (`r filter(excl$test$no_response, variable == 'injury_severity_ais')$plot_cap`), significantly lower rates of hospitalization (`r filter(excl$test$no_response, variable == 'hospitalization')$plot_cap`) and surgery (`r filter(excl$test$no_response, variable == 'surgery_done')$plot_cap`), as well as significantly lower surgery complexity reflected by surgical diagnosis numbers (`r filter(excl$test$no_response, variable == 'surgery_complexity')$plot_cap`).</span>
<span custom-style = 'revision'>Effect sizes of those differences were small (__Supplementary Table S4__).</span> 
The survey participants excluded from the analysis due to incomplete psychometric data had significantly 
lower income <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'household_income_class')$plot_cap`)</span>, 
less severe injury <span custom-style = 'revision'>grade (`r filter(excl$test$missing, variable == 'injury_severity_ais')$plot_cap`)</span>, 
more frequent injuries of the upper limbs <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'injury_upper_limbs')$plot_cap`)</span>, 
less frequently required hospitalization <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'hospitalization')$plot_cap`)</span> or 
surgery <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'surgery_done')$plot_cap`)</span>, and 
suffered less frequently from persistent physical health consequences of the accident than the analyzed participants <span custom-style = 'revision'>(`r filter(excl$test$missing, variable == 'accident_aftermath')$plot_cap`)</span>. 
Effect sizes of the differences between the included and excluded participants were weak <span custom-style = 'revision'>(__Supplementary Table S5__).</span>
